---
title: "Dpa_data_preprocessing"
author: "Bharath Srinivas SR"
date: "2025-04-13"
output: html_document
---

Note: See for Route Numbers:

```{r}
# Importing the CSV file
crash_data <- read.csv("crash_22.csv", header = TRUE)


# View the first few rows of the dataset
head(crash_data)

dim(crash_data)

```

We will now take a count on the missing values present in the dataset

```{r}

missing_per_column <- colSums(is.na(crash_data))
print(missing_per_column)

```
Given that the AlignmentCode column has almost all values missing (298,324 out of 298,431 rows), and ClassOfTrafficwayCode has all missing ,we will drop them both.


```{r}
# Count the number of "(UNK)" values in each column
unknown_counts <- sapply(crash_data, function(col) sum(col == "(UNK)", na.rm = TRUE))

# Filter only columns that have at least one "(UNK)"
unknown_counts <- unknown_counts[unknown_counts > 0]

# Print the result
print(unknown_counts)
```
NumberOfLanes, RoadAlignment,RouteNumber and AccessControl can be removed. Flow condition, Roadway functional class shouldn't be removed because important for analysisng dataset.

```{r}

library(dplyr)

# Drop both ClassOfTrafficwayCode and AlignmentCode columns
clean_data <- crash_data %>% select(-ClassOfTrafficwayCode, -AlignmentCode,-NumberOfLanes, -RoadAlignment,-AccessControl,-RouteNumber)
dim(clean_data)
```

```{r}
missing_per_column <- colSums(is.na(clean_data))
print(missing_per_column)
```

```{r}
# Count the number of "(UNK)" values in each column
unknown_counts <- sapply(clean_data, function(col) sum(col == "(UNK)", na.rm = TRUE))

# Filter only columns that have at least one "(UNK)"
unknown_counts <- unknown_counts[unknown_counts > 0]

# Print the result
print(unknown_counts)
```

0.0017 % of the dataset 517- we drop rows with (UNK) ClassOfTrafficway,UrbanRural,DidInvolveSecondaryCrash,WorkZone

We will now drop rows( apart from RouteNumber ) which have minimal missing values

columns with relatively few missing values include:

RoadwayFunctionalClassCode (198 missing values)

DamagedProperty1 (1 missing value)

DamagedProperty2 (5 missing values)

TSCrashLatitude and TSCrashLongitude (533 missing values each)

```{r}
clean_data <- clean_data %>%
  filter(
    ClassOfTrafficway != "(UNK)",
    UrbanRural != "(UNK)",
    DidInvolveSecondaryCrash != "(UNK)",
    WorkZone != "(UNK)",
    AgencyReportNumber!="(UNK)"
  )

clean_data <- clean_data %>%
  filter(
    !is.na(TSCrashLatitude),
    !is.na(TSCrashLongitude),
    !is.na(DamagedProperty1),
    !is.na(AgencyReportYear)
  )

# Count the number of "(UNK)" values in each column
unknown_counts <- sapply(clean_data, function(col) sum(col == "(UNK)", na.rm = TRUE))

# Filter only columns that have at least one "(UNK)"
unknown_counts <- unknown_counts[unknown_counts > 0]

# Print the result
print(unknown_counts)
```

CAN'T DIRECTLY DROP ROADWAY FUNCTIONAL CLASSES:
```{r}
library(ggplot2)
# Plot distribution of Roadway Functional Classes
clean_data %>%
  count(RoadwayFunctionalClass, sort = TRUE) %>%
  mutate(RoadwayFunctionalClass = factor(RoadwayFunctionalClass, levels = rev(RoadwayFunctionalClass))) %>%
  ggplot(aes(x = RoadwayFunctionalClass, y = n)) +
  geom_bar(stat = "identity", fill = "darkcyan") +
  labs(title = "Crash Counts by Roadway Functional Class",
       x = "Roadway Functional Class",
       y = "Number of Crashes") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = n), hjust = -0.1, size = 4) +
  coord_flip()

```


Interstate- Major highways/freeways (I-55, I-90, etc.), designed for high-speed travel
Principal Arterial- Key roads connecting major centers; high traffic volumes
Minor Arterial- Connects smaller cities or neighborhoods; moderate traffic
Major Collector - Gathers traffic from local streets to arterial roads
Minor Collector	- Feeds into major collectors; lower speeds and traffic
Local - or Local Road	Residential or small local access roads
(UNK)	- Unknown or not classified — often used when info is missing

They haven't recorded it properly there are too many (UNK).

Trying to solve this problem by looking a the other columns.
```{r}
unique(clean_data$Toll)
unique(clean_data$UrbanRural)
unique(clean_data$FlowCondition)
```

So we make the following assumptions using the other columns to get a better plot:
We used Toll == "Illinois Toll Highway Com" and gave it "Interstate" because this directly represents Illinois’ toll-operated interstate highways.
We used Toll == "No Toll" or "Not Toll" and gave it "Principal Arterial" because non-toll roads are generally public through-routes, often designed to handle moderate to high traffic regardless of area type.
We used UrbanRural == "URBAN" and FlowCondition == "Free Flow" and gave it "Principal Arterial" because fast-moving traffic in cities suggests high-capacity arterial roads.
We used UrbanRural == "URBAN" and FlowCondition == "Slow" and gave it "Major Collector" because slow traffic in urban areas often indicates moderately busy collector roads.
We used UrbanRural == "URBAN" and FlowCondition == "Stopped" and gave it "Minor Arterial" because stopped traffic in cities often occurs on important but lower-tier roads.
We used UrbanRural == "RURAL" and FlowCondition == "Stopped" and gave it "Major Collector" because heavy use in rural areas points to key connecting roads.
We used UrbanRural == "RURAL" and FlowCondition == "Slow" and gave it "Minor Collector" because these roads tend to be small, secondary routes in rural regions.
We used FlowCondition == "No" and gave it "Local" because the absence of traffic data likely indicates low-traffic local roads.
We used TRUE and gave it "(UNK)" because none of the defined conditions matched, so we conservatively left it unknown.

```{r}
# Plotting again
library(dplyr)

clean_data <- clean_data %>%
  mutate(RoadwayFunctionalClass = case_when(
    RoadwayFunctionalClass != "(UNK)" ~ RoadwayFunctionalClass,
    Toll == "Illinois Toll Highway Com" ~ "Interstate",
    Toll == "No Toll" | Toll=="Not Toll" ~ "Other Principal Arterial",
    UrbanRural == "Urban" & FlowCondition == "Free Flow" ~ "Principal Arterial",
    UrbanRural == "Urban" & FlowCondition == "Slow" ~ "Major Collector",
    UrbanRural == "Urban" & (FlowCondition == "Stopped" |FlowCondition =="Slow") ~ "Minor Arterial",
    UrbanRural == "Rural" & FlowCondition == "Stopped" ~ "Major Collector",
    UrbanRural == "Rural" & FlowCondition == "Slow" ~ "Minor Collector",
    TRUE ~ "(UNK)"
  ))

clean_data %>%
  count(RoadwayFunctionalClass, sort = TRUE) %>%
  mutate(RoadwayFunctionalClass = factor(RoadwayFunctionalClass, levels = rev(RoadwayFunctionalClass))) %>%
  ggplot(aes(x = RoadwayFunctionalClass, y = n)) +
  geom_bar(stat = "identity", fill = "darkcyan") +
  labs(title = "Crash Counts by Roadway Functional Class",
       x = "Roadway Functional Class",
       y = "Number of Crashes") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = n), hjust = -0.1, size = 4) +
  coord_flip()
```



```{r}
library(dplyr)

# Fill missing Cause1 and Cause2 from each other
clean_data <- clean_data %>%
  mutate(
    Cause1 = ifelse((is.na(Cause1) | Cause1 == "") & !is.na(Cause2) & Cause2 != "", Cause2, Cause1),
    Cause2 = ifelse((is.na(Cause2) | Cause2 == "") & !is.na(Cause1) & Cause1 != "", Cause1, Cause2)
  )

# Get top 5 causes
cause1_top <- clean_data %>%
  count(Cause1, sort = TRUE) %>%
  slice_max(n, n = 5)

cause2_top <- clean_data %>%
  count(Cause2, sort = TRUE) %>%
  slice_max(n, n = 5)

# Set plotting layout: one below the other
par(mfrow = c(2, 1), mar = c(5, 12, 4, 2))

# Plot 1: Cause1
barplot(cause1_top$n,
        names.arg = cause1_top$Cause1,
        main = "Top 5 Causes (Cause1)",
        col = "tomato",
        xlab = "Number of Crashes",
        horiz = TRUE,
        las = 1)

# Plot 2: Cause2
barplot(cause2_top$n,
        names.arg = cause2_top$Cause2,
        main = "Top 5 Causes (Cause2)",
        col = "tomato",
        xlab = "Number of Crashes",
        horiz = TRUE,
        las = 1)

```

```{r}
sum(clean_data$Cause2 == "(N/A)", na.rm = TRUE)
```

Too many NA values so this column can be dropped.
```{r}
clean_data <- clean_data %>% select(-Cause2)
dim(clean_data)
```

Let us start with EDA now.(havent dealt with routeNumber coulumn yet)


## Summary Statistics 

```{r}
summary(clean_data)

```

```{r}

str(clean_data)


```

### Missing Data analysis

```{r}
colSums(is.na(clean_data))

```

### Distribution Analysis

####Numerical Variables

Let us understand the distribution of numerical features.

We will plot histograms for numerical variables to observe their distributions:

```{r}
#increase size
#install.packages("reshape2")
library(ggplot2)
library(reshape2)

# Melt the dataset for plotting
numeric_cols <- sapply(clean_data, is.numeric)
melted_data <- melt(clean_data[, numeric_cols])

# Create histogram plots with adjusted layout
ggplot(melted_data, aes(value)) +
  geom_histogram(fill = "steelblue", color = "black") +
  facet_wrap(~variable, ncol = 2, scales = "free") +  # Set 2 columns
  labs(title = "Distribution of Numerical Variables", x = "Value", y = "Count")
```
### Correlation Analysis

####Numerical Variables

#####Let us identify relationships between numerical features and the target variable (CrashSeverityCd).



### Strong correlation numerical columns with CrashInjurySeverity:
```{r}
library(dplyr)

# Step 1: Filter numeric columns
numeric_cols <- sapply(clean_data, is.numeric)
numeric_data <- clean_data[, numeric_cols]

# Step 2: Remove columns with zero variance
numeric_data <- numeric_data[, sapply(numeric_data, function(col) sd(col, na.rm = TRUE) != 0)]

# Step 3: Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Step 4: Identify columns with correlation > 0.6 with any other column (excluding self-correlation)
high_corr_columns <- colnames(cor_matrix)[
  apply(cor_matrix, 2, function(x) any(abs(x[!is.na(x) & names(x) != names(x)[which.max(abs(x))]]) > 0.6))
]

# Step 5: Print the selected column names
cat("Columns with at least one strong correlation (|r| > 0.6) with any other column:\n")
print(high_corr_columns)

# Step 6: Get correlation with CrashInjurySeverity only (if it exists)
if ("CrashInjurySeverity" %in% rownames(cor_matrix)) {
  target_cor <- cor_matrix["CrashInjurySeverity", ]
  
  # Remove self-correlation and filter those above threshold
  high_corr_with_target <- target_cor[abs(target_cor) > 0.6 & names(target_cor) != "CrashInjurySeverity"]
  
  # Print
  cat("\nNumeric columns with |correlation| > 0.6 with CrashInjurySeverity:\n")
  print(high_corr_with_target)
} else {
  cat("\nCrashInjurySeverity is not in the correlation matrix.\n")
}
```
### strong Correlation categorical columns with CrashInjurySeverity:
```{r}
library(dplyr)

# Identify categorical columns
categorical_cols <- sapply(clean_data, function(col) is.character(col) || is.factor(col))

# Run Fisher’s test if needed, otherwise use Chi-squared
significant_vars <- lapply(names(clean_data)[categorical_cols], function(var) {
  if (var != "CrashInjurySeverity") {
    tbl <- table(clean_data[[var]], clean_data$CrashInjurySeverity)
    test <- tryCatch({
      if (all(dim(tbl) <= 5)) {
        fisher.test(tbl)
      } else {
        chisq.test(tbl)
      }
    }, error = function(e) return(NULL))
    
    if (!is.null(test) && inherits(test, "htest") && test$p.value < 0.05) {
      return(var)
    }
  }
  return(NULL)
})

significant_vars <- unlist(significant_vars)
cat("✅ Reliable variables with p < 0.05 (Chi-sq or Fisher’s Exact Test):\n")
print(significant_vars)
```

#Strong correlation numerical columns with CrashSeverity:
```{r}
library(dplyr)

# Step 1: Filter numeric columns
numeric_cols <- sapply(clean_data, is.numeric)
numeric_data <- clean_data[, numeric_cols]

# Step 2: Remove columns with zero variance
numeric_data <- numeric_data[, sapply(numeric_data, function(col) sd(col, na.rm = TRUE) != 0)]

# Step 3: Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Step 4: Identify columns with correlation > 0.6 with any other column (excluding self-correlation)
high_corr_columns <- colnames(cor_matrix)[
  apply(cor_matrix, 2, function(x) any(abs(x[!is.na(x) & names(x) != names(x)[which.max(abs(x))]]) > 0.6))
]

# Step 5: Print the selected column names
cat("Columns with at least one strong correlation (|r| > 0.6) with any other column:\n")
print(high_corr_columns)

# Step 6: Get correlation with CrashInjurySeverity only (if it exists)
if ("CrashInjurySeverity" %in% rownames(cor_matrix)) {
  target_cor <- cor_matrix["CrashSeverity", ]
  
  # Remove self-correlation and filter those above threshold
  high_corr_with_target <- target_cor[abs(target_cor) > 0.6 & names(target_cor) != "CrashSeverity"]
  
  # Print
  cat("\nNumeric columns with |correlation| > 0.6 with CrashSeverity:\n")
  print(high_corr_with_target)
} else {
  cat("\nCrashSeverity is not in the correlation matrix.\n")
}
```
#Strong correlation categorical columns with CrashSeverity:
```{r}
library(dplyr)

# Identify categorical columns
categorical_cols <- sapply(clean_data, function(col) is.character(col) || is.factor(col))

# Run Fisher’s test if needed, otherwise use Chi-squared
significant_vars <- lapply(names(clean_data)[categorical_cols], function(var) {
  if (var != "CrashSeverity") {
    tbl <- table(clean_data[[var]], clean_data$CrashSeverity)
    test <- tryCatch({
      if (all(dim(tbl) <= 5)) {
        fisher.test(tbl)
      } else {
        chisq.test(tbl)
      }
    }, error = function(e) return(NULL))
    
    if (!is.null(test) && inherits(test, "htest") && test$p.value < 0.05) {
      return(var)
    }
  }
  return(NULL)
})

significant_vars <- unlist(significant_vars)
cat("Reliable variables with p < 0.05 (Chi-sq or Fisher’s Exact Test):\n")
print(significant_vars)
```

#Combining all the columns:
```{r}
selected_columns <- c(
  "X", "Y", "ICN", "CrashID", "CrashMonth", "CityCode", "CityClassCode", 
  "TotalInjured", "BInjuries", "CInjuries", "TSCrashCoordinateX", "TSCrashLatitude", "TSCrashLongitude", 
  "CrashSeverityCd", "CrashInjurySeverity", "CrashSeverity",
  
  "Township", "RailroadCrossingNumber", "TimeOfCrash", "CrashDate", 
  "CrashReportCounty", "DayOfWeek", "TypeOfFirstCrash", "CityName", 
  "CityClass", "ClassOfTrafficway", "Cause1", "TrafficControlDevice", 
  "TrafficControlDeviceCond", "RoadSurfaceCond", "RoadDefects", 
  "LightingCond", "WeatherCond", "TrafficwayDescrip", 
  "RoadwayFunctionalClass", "DamagedProperty1", "DamagedProperty2", 
  "AgencyReportNumber"
)

# Create a filtered dataset
clean_data_filter <- clean_data[, selected_columns]
head(clean_data_filter)
dim(clean_data_filter)
```

Note: Remove BInjuries,CInjuries why- one hot for binjury can be done in model development
TScrashCoordinateX,TSCrashLatitude,TSCrashLongitude- because highly correlate with "X", "Y".
Removing ICN because it is unique identification for each crash report so plays the same role as CrashID.
Removing DamagedProperty2 because similar to Damaged property1 and most values are empty
Removing RailroadCrossingNumber mostly values are N/A.
AgencyReportNumber- Not much important as we are already taking county into account.

```{r}
# Columns to remove
selected_columns <- c("BInjuries", "TSCrashCoordinateX", "TSCrashLatitude", "TSCrashLongitude", "ICN","DamagedProperty2","CInjuries","RailroadCrossingNumber","AgencyReportNumber")

# Remove them from the dataframe
clean_data_filter <- clean_data_filter[, setdiff(names(clean_data_filter), selected_columns)]

# Check the updated dimensions or preview
dim(clean_data_filter)
head(clean_data_filter)
```
```{r}
#write.csv(clean_data_filter, "clean_data_filter.csv", row.names = FALSE)
```


####Numerical Features vs Target Variable

We will explore how numerical features relate to crash severity.



Will use boxplots or scatterplots for key numerical features  vs CrashSeverity:

```{r}
ggplot(clean_data_filter, aes(x = CrashSeverity, y = CrashMonth)) +
  geom_boxplot() +
  labs(title = " Crash Severity by Months")
```

#### Categorical Features vs Target Variable
we will analyze how categorical features influence crash severity.

```{r}
ggplot(clean_data_filter, aes(x = CrashSeverity, fill =WeatherCond)) +
  geom_bar(position = "dodge") +
  labs(title = "Weather Conditions vs Crash Severity")

```
```{r}
ggplot(clean_data_filter, aes(x = CrashSeverity, fill =RoadSurfaceCond)) +
  geom_bar(position = "dodge") +
  labs(title = "Road Conditions vs Crash Severity")
```


```{r}
colnames(clean_data_filter)
```
#### Feature importance

we can do some preliminary steps to have an idea on some of the important predictors for crash severity


```{r}
# Select numeric columns
numeric_vars <- clean_data_filter %>% select(where(is.numeric))

# Correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Visualize
library(corrplot)
corrplot(cor_matrix, method = "color", tl.cex = 0.7)
```


```{r}
library(lubridate)

# If the column is still character
clean_data_filter$CrashDate <- as.character(clean_data_filter$CrashDate)

# Parse the datetime with timezone
clean_data_filter$CrashDate <- ymd_hms(clean_data_filter$CrashDate, tz = "UTC")

# Split into Date and Time
clean_data_filter$CrashDate_only <- as.Date(clean_data_filter$CrashDate)
clean_data_filter$CrashTime_only <- format(clean_data_filter$CrashDate, "%H:%M:%S")
# Step 1: Replace CrashDate with CrashDate_only
clean_data_filter$CrashDate <- clean_data_filter$CrashDate_only

# Step 2: Remove CrashDate_only and CrashTime_only
clean_data_filter$CrashDate_only <- NULL
clean_data_filter$CrashTime_only <- NULL
```

```{r}
library(dplyr)

clean_data_filter <- clean_data_filter %>%
  arrange(CrashDate)
```



```{r}
mappings <- list(
  CrashInjurySeverity = levels(as.factor(clean_data_filter$CrashInjurySeverity)),
  RoadwayFunctionalClass = levels(as.factor(clean_data_filter$RoadwayFunctionalClass)),
  WeatherCond = levels(as.factor(clean_data_filter$WeatherCond)),
  LightingCond = levels(as.factor(clean_data_filter$LightingCond)),
  RoadSurfaceCond = levels(as.factor(clean_data_filter$RoadSurfaceCond))
)

print(mappings$CrashInjurySeverity)
print(mappings$RoadwayFunctionalClass)
print(mappings$LightingCond)
print(mappings$WeatherCond)
print(mappings$RoadSurfaceCond)
```
getting rid of the columns now after
```{r}
# Label encode and create new columns
clean_data_filter$CrashInjurySeverityCond <- as.numeric(as.factor(clean_data_filter$CrashInjurySeverity))
clean_data_filter$RoadwayFunctionalClassCond <- as.numeric(as.factor(clean_data_filter$RoadwayFunctionalClass))
clean_data_filter$WeatherCondCond <- as.numeric(as.factor(clean_data_filter$WeatherCond))
clean_data_filter$LightingCondCond <- as.numeric(as.factor(clean_data_filter$LightingCond))
clean_data_filter$RoadSurfaceCondCond <- as.numeric(as.factor(clean_data_filter$RoadSurfaceCond))
```

```{r}
# Ensure CrashDate is of Date type
clean_data_filter$CrashDate <- as.Date(clean_data_filter$CrashDate)

# Filter rows where CrashDate is on or before 30 June 2022
clean_data <- clean_data_filter %>%
  filter(CrashDate <= as.Date("2022-06-30"))

# Save to CSV
#write.csv(clean_data_before_cutoff, "clean_dataBeforeFS_new.csv", row.names = FALSE)
```

```{r}
# Ensure CrashDate is of Date type
dim(clean_data)
head(clean_data)
```
##Feature Selection:
```{r}
# Convert Township to numeric using factor encoding
clean_data_filter$Township <- as.numeric(as.factor(clean_data_filter$Township))

```

#only works on Numerical columns the best.
```{r}
# Take first 50,000 rows as before
data <- clean_data

# Add numeric response
data$Y <- as.numeric(as.factor(data$CrashSeverity))

# Drop the original target and non-numeric columns
data$CrashSeverity <- NULL

# Keep only numeric columns (including Y)
numeric_data <- data[, sapply(data, is.numeric)]
```

#Forward

```{r}
library(leaps)

regfit.fwd <- regsubsets(Y ~ ., data = numeric_data, nvmax = 8, method = "forward")
summary_fwd <- summary(regfit.fwd)
```

```{r}
best_features <- summary_fwd$which[which.max(summary_fwd$adjr2), ]
selected_vars <- names(best_features[best_features == TRUE])
selected_vars <- selected_vars[selected_vars != "(Intercept)"]

print(selected_vars)
```
#Backward Selection:


```{r}
library(leaps)

# Run backward selection
regfit.bwd <- regsubsets(Y ~ ., data = numeric_data, nvmax = 8, method = "backward")

# Get summary
summary_bwd <- summary(regfit.bwd)
```

```{r}
best_bwd <- summary_bwd$which[which.max(summary_bwd$adjr2), ]
selected_bwd_vars <- names(best_bwd[best_bwd == TRUE])
selected_bwd_vars <- selected_bwd_vars[selected_bwd_vars != "(Intercept)"]
print(selected_bwd_vars)
```

Gave same results for both, so moving forward with these 6 numerical columns and all the rest categorical columns.

# ridge Lasso:
```{r}
# Load required libraries
library(glmnet)

# Step 2: Prepare x (features) and y (response)
x <- as.matrix(numeric_data[, setdiff(names(numeric_data), "Y")])
y <- numeric_data$Y
```

#ridge:
```{r}
# Fit Ridge with cross-validation
ridge_cv <- cv.glmnet(x, y, alpha = 0)

# Coefficients at best lambda
ridge_coef <- coef(ridge_cv, s = "lambda.min")
print("Ridge selected features:")
print(ridge_coef[ridge_coef[,1] != 0, , drop=FALSE])
```
Based on Ridge regression coefficients, the top 8 selected numeric features are: CrashSeverityCd, CrashInjurySeverity, TotalInjured, CityClassCode, WeatherCondCond, RoadSurfaceCondCond, CrashMonth and RoadwayFunctionalClassCond.

#Lasso:
```{r}
# Fit Lasso with cross-validation
lasso_cv <- cv.glmnet(x, y, alpha = 1)

# Coefficients at best lambda
lasso_coef <- coef(lasso_cv, s = "lambda.min")
print("Lasso selected features:")
print(lasso_coef[lasso_coef[,1] != 0, , drop=FALSE])
```
Similar to previous.
So taking only the top 8 features from the numerical column and also crashID as it is imp

#new dataset:
```{r}
# === Step 1: Define the numeric columns you want to keep ===
selected_numeric_cols <- c(
  "CrashSeverityCd",
  "CrashInjurySeverity",
  "TotalInjured",
  "CityClassCode",
  "CrashMonth"
)

# === Step 2: Get all categorical columns from original data ===
categorical_cols <- names(clean_data_filter)[sapply(clean_data_filter, function(col) is.factor(col) || is.character(col))]


# === Step 3: Combine numeric and categorical columns ===
final_columns <- union(selected_numeric_cols, categorical_cols)

# === Step 4: Remove unwanted encoded columns ===
columns_to_exclude <- c("RoadwayFunctionalClassCond", "WeatherCondCond", "LightingCondCond", "RoadSurfaceCondCond")

final_columns <- setdiff(final_columns, columns_to_exclude)

# === Step 5: Create final dataframe ===

final_df <- clean_data[, final_columns]

# === Step 6: Check structure ===
str(final_df)
dim(final_df)

```

```{r}
# Save the final dataframe to a CSV file

write.csv(final_df, "final_dataset.csv", row.names = FALSE)
```

EDA:
```{r}
# Step 1: Create labels for months
month_labels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Step 2: Prepare data tables
month_counts <- table(final_df$CrashMonth)
names(month_counts) <- month_labels[as.numeric(names(month_counts))]

road_counts <- table(final_df$RoadSurfaceCond)
weather_counts <- table(final_df$WeatherCond)

# Step 3: Set layout for 3 plots side-by-side
par(mfrow = c(1, 3), mar = c(6, 5, 4, 2))

# Plot 1: Crash by Month
barplot(month_counts,
        main = "Crash Counts by Month",
        col = "steelblue",
        ylab = "Number of Crashes",
        las = 2)

# Plot 2: Crash by Road Surface Condition
barplot(road_counts,
        main = "By Road Surface Condition",
        col = "darkgreen",
        ylab = "Number of Crashes",
        las = 2)

# Plot 3: Crash by Weather Condition
barplot(weather_counts,
        main = "By Weather Condition",
        col = "darkorange",
        ylab = "Number of Crashes",
        las = 2)

```


Most accidents happened in the month of October and are higher in the months of Nov,dec mostly because of Snow?(Why?), most of the time the codition was clear.

```{r}
library(ggplot2)
library(lubridate)
library(dplyr)

# Ensure TimeOfCrash is properly parsed as a time object
final_df <- final_df %>%
  mutate(TimeOfCrash = hms::as_hms(TimeOfCrash)) %>%  # <--- FIX: convert to time
  mutate(TimeCategory = case_when(
    hour(TimeOfCrash) >= 6  & hour(TimeOfCrash) < 12 ~ "Morning",
    hour(TimeOfCrash) >= 12 & hour(TimeOfCrash) < 15 ~ "Afternoon",
    hour(TimeOfCrash) >= 15 & hour(TimeOfCrash) < 18 ~ "Evening",
    TRUE ~ "Night"
  ))


# Plot 1: Time of Day
barplot(time_counts,
        main = "Crash Counts by Time of Day",
        col = "darkorange",
        ylab = "Number of Crashes",
        las = 2)

```
It's surprising to se most of the accidents in Illinois happen in Morning/daylight.

```{r}
#for Type of First accident
library(ggplot2)
library(dplyr)

# Plot: Type of First Crash
final_df %>%
  count(TypeOfFirstCrash, sort = TRUE) %>%
  mutate(TypeOfFirstCrash = factor(TypeOfFirstCrash, levels = rev(unique(TypeOfFirstCrash)))) %>%
  ggplot(aes(x = TypeOfFirstCrash, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Crash Counts by Type of First Crash",
       x = "Type of First Crash",
       y = "Number of Crashes") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = n), hjust = -0.1, size = 3) +
  coord_flip()  # Makes it horizontal for readability

```

```{r}
#top 10 Cities with most accidents
library(dplyr)
library(ggplot2)

# Step 1: Count crashes by city and keep top 10
top_cities <- final_df %>%
  count(CityName, sort = TRUE) %>%
  slice_max(n, n = 10)

# Step 2: Plot
top_cities %>%
  mutate(CityName = factor(CityName, levels = rev(CityName))) %>%
  ggplot(aes(x = CityName, y = n)) +
  geom_bar(stat = "identity", fill = "firebrick") +
  labs(title = "Top 10 Cities with Most Crashes",
       x = "City",
       y = "Number of Crashes") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = n), hjust = -0.1, size = 4) +
  coord_flip()  # Horizontal bar chart for readability
```
Chicago has the mosr accidents. with a very high rate when compared to the other cities


### After completing EDA, we know a lot of details of our dataset like:

Key distributions (numerical and categorical).

Missing data patterns and decisions made.

Correlations between features and the target variable.

Relationships between predictors and crash severity

