---
title: "DPA Model Development"
author: "Jacob Callahan"
date: "2025-04-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(caret)
library(nnet)
library(DBI)
library(RSQLite)
library(dplyr)
library(rsample)
source("utils.R")
library(e1071)
library(xgboost)
library(synthpop)
library(ranger)
library(MLmetrics)
library(doParallel)
```

## Loading Data

```{r}
con <- dbConnect(SQLite(), "crashes.db")

# Get the column names
query <- "
SELECT *
FROM crashes;"
# ORDER BY RANDOM()
# LIMIT 500000;"

orig_data = dbGetQuery(con, query)

# Leaves just CrashSeverity target Column
orig_data = orig_data[,!(colnames(orig_data) %in% c("TotalInjured","CrashInjurySeverity"))]

# Close connection
dbDisconnect(con)

orig_data = orig_data[,!colnames(orig_data) %in% c("HourOfCrash","CrashYr")]
                              
for (col in names(orig_data)) {
  orig_data[[col]] <- factor(orig_data[[col]])
}

data = orig_data

table(data$CrashSeverity)

```

## Developing Baseline Model

Our baseline model will be a simple multinomial Logistic Regression. Our response variable will be *CrashSeverity*. Our sole predictor will be *WeatherCond*. The three responses we will be predicting can be Fatal, Injury, or Property Damage.


### Creating train test split

```{r}
set.seed(5)
split = get_train_test(data, "CrashSeverity", data_slice_pct = .05, seed = 1)

train = split$train
test = split$test

```

### Baseline Model Training

```{r}

k_folds <- 5
folds <- createFolds(train$CrashSeverity, k = k_folds, list = TRUE)

# Store metrics
f1s = data.frame(Macro_F1 = numeric(), Weighted_F1 = numeric())

all_probs <- list()
all_actuals <- c()
all_preds <- c()

for (i in 1:k_folds) {
  val_idx <- folds[[i]]
  train_cv <- train[-val_idx, ]
  val_cv <- train[val_idx, ]

  # Fit model
  model_cv <- multinom(CrashSeverity ~ WeatherCond, data = train_cv)

  # Predictions
  probs_cv <- predict(model_cv, val_cv, type = "prob")
  class_cv <- predict(model_cv, val_cv, type = "class")

  # F1 score
  f1_scores = get_f1_vals(val_cv$CrashSeverity, class_cv)
  
  f1s = rbind(f1s, f1_scores)

  # Collect probs and actuals for ROC
  all_probs <- append(all_probs, list(probs_cv))
  all_actuals <- c(all_actuals, as.character(val_cv$CrashSeverity))
  all_preds <- c(all_preds, as.character(class_cv))
}

colMeans(f1s)

probs_matrix <- do.call(rbind, all_probs)

# Plot ROC
plot_roc_OvR(probs_matrix, all_actuals)


table(predicted = all_preds, actual = all_actuals)


```

Clearly we can see that the baseline model is not very effective at predicting anything but Property Damage. 

## Naive Bayes

### Creating train test split

```{r}
set.seed(5)
split = get_train_test(data, "CrashSeverity", data_slice_pct = .05, seed = 1)

train = split$train
test = split$test

```

### Cross Validation Naive Bayes

```{r}
library(dplyr)
library(e1071)


k_folds <- 5
folds <- createFolds(train$CrashSeverity, k = k_folds, list = TRUE)

# Store metrics
f1s = data.frame(Macro_F1 = numeric(), Weighted_F1 = numeric())

all_probs <- list()
all_actuals <- c()
all_preds <- c()

for (i in 1:k_folds) {
  val_idx <- folds[[i]]
  train_cv <- train[-val_idx, ]
  val_cv <- train[val_idx, ]

  # Fit model
  model_cv = naiveBayes(CrashSeverity ~., data = train, laplace = 0)

  # Predictions
  probs_cv <- predict(model_cv, val_cv, type = "raw")
  class_cv <- predict(model_cv, val_cv, type = "class")

  # F1 score
  f1_scores = get_f1_vals(val_cv$CrashSeverity, class_cv)
  
  f1s = rbind(f1s, f1_scores)

  # Collect probs and actuals for ROC
  all_probs <- append(all_probs, list(probs_cv))
  all_actuals <- c(all_actuals, as.character(val_cv$CrashSeverity))
  all_preds <- c(all_preds, as.character(class_cv))
}

colMeans(f1s)

probs_matrix <- do.call(rbind, all_probs)

# Plot ROC
plot_roc_OvR(probs_matrix, all_actuals)


table(predicted = all_preds, actual = all_actuals)



```



## Adding predictors to baseline model

Our initial model was only partially effective as we just had a single predictor. Now we will add all predictors for a multiple multinomial logistic regression.

Using library "nnet" to conduct multinomial logistic regression. 

### Creating train test split

```{r}
set.seed(5)
split = get_train_test(data, "CrashSeverity", data_slice_pct = .05, seed = 1)

train = split$train
test = split$test

```

### Multiple Logistic Regression

```{r}
library(nnet)

k_folds <- 5
folds <- createFolds(train$CrashSeverity, k = k_folds, list = TRUE)

# Store metrics
f1s = data.frame(Macro_F1 = numeric(), Weighted_F1 = numeric())

all_probs <- list()
all_actuals <- c()
all_preds <- c()

for (i in 1:k_folds) {
  val_idx <- folds[[i]]
  train_cv <- train[-val_idx, ]
  val_cv <- train[val_idx, ]

  # Fit model
  model_cv <- multinom(CrashSeverity ~ ., data = train_cv)

  # Predictions
  probs_cv <- predict(model_cv, val_cv, type = "prob")
  class_cv <- predict(model_cv, val_cv, type = "class")

  # F1 score
  f1_scores = get_f1_vals(val_cv$CrashSeverity, class_cv)
  
  f1s = rbind(f1s, f1_scores)

  # Collect probs and actuals for ROC
  all_probs <- append(all_probs, list(probs_cv))
  all_actuals <- c(all_actuals, as.character(val_cv$CrashSeverity))
  all_preds <- c(all_preds, as.character(class_cv))
}

colMeans(f1s)

probs_matrix <- do.call(rbind, all_probs)

# Plot ROC
plot_roc_OvR(probs_matrix, all_actuals)


table(predicted = all_preds, actual = all_actuals)


```

```{r}
model.log <- multinom(CrashSeverity ~ ., data = train)

coefs <- summary(model.log)$coefficients
std_err <- summary(model.log)$standard.errors

z_values <- coefs / std_err
p_values <- 2 * (1 - pnorm(abs(z_values)))  

```


## SVM

### Train/Test Split

```{r}

split = get_train_test(orig_data, "CrashSeverity",data_slice_pct = .01, train_pct = .5)
train = split$train
test = split$test

ctrl <- trainControl(method = "cv", number = 5)  # 5-fold CV

n_cores <- parallel::detectCores()

```


### Linear Cross-Validation and Hyperparameter tuning

```{r}


cl <- makePSOCKcluster(12)  # Leave 1 core free to keep system responsive
registerDoParallel(cl)

grid <- expand.grid(
  cost = c(0.1, 1, 10)        
)

svm.cv.linear <- train(
  CrashSeverity ~ ., 
  data = train,
  trControl = ctrl,
  method = "svmLinear2",
  tuneGrid = grid
)

svm.cv.linear

stopCluster(cl)
registerDoSEQ()


```

Accuracy was used to select the optimal model using the largest value.
The final value used for the model was cost = 0.1.

### Training Linear Best Model

```{r}

svm.linear <- train(
  CrashSeverity ~ ., 
  data = train,
  method = "svmLinear",
  tuneGrid = data.frame(C = .1)                # your chosen cost value
)

yhats = predict(svm.linear, newdata = test)

table(predicted = yhats, actual = test[,1])

get_f1_vals(test[,1],yhats)

```



### Radial Cross-Validation and Hyperparameter tuning

```{r}

cl <- makePSOCKcluster(n_cores - 1)  # Leave 1 core free to keep system responsive
registerDoParallel(cl)

grid <- expand.grid(
  C = c(0.1, 1, 10),        
  sigma = c(0.0001, 0.001, 0.01)
)

svm.cv.rad <- train(
  CrashSeverity ~ ., 
  data = train,
  trControl = ctrl,
  method = "svmRadial",
  tuneGrid = grid
)

svm.cv.rad

stopCluster(cl)
registerDoSEQ()

```
Accuracy was used to select the optimal model using the largest value.
The final values used for the model were sigma = 0.01 and C = 1

### Training best Radial Model

```{r}

cl <- makePSOCKcluster(12)  # Leave 1 core free to keep system responsive
registerDoParallel(cl)

svm.radial <- train(
  CrashSeverity ~ ., 
  data = train,
  method = "svmRadial",
  tuneGrid = data.frame(C = 1, sigma = 0.01)
  )

yhats = predict(svm.radial, newdata = test)

table(predicted = yhats, actual = test[,1])

get_f1_vals(test[,1],yhats)

stopCluster(cl)
registerDoSEQ()

```




### Polynomial Cross-Validation and Hyperparameter tuning

```{r}

grid <- expand.grid(
  degree = c(2, 3, 4),   # degree of the polynomial
  scale = c(0.001, 0.01, 0.1),  # scaling factor
  C = c(0.1, 1, 10)      # regularization parameter
)

svm.cv.poly <- train(
  CrashSeverity ~ ., 
  data = train,
  trControl = ctrl,
  method = "svmPoly",
  tuneGrid = grid
)

svm.cv.poly



```

Accuracy was used to select the optimal model using the largest value.
The final values used for the model were degree = 2, scale = 0.1 and C = 1.

### Training best Polynomial Model

```{r}

cl <- makePSOCKcluster(12)
registerDoParallel(cl)


svm.poly <- train(
  CrashSeverity ~ ., 
  data = train,
  method = "svmPoly",
  tuneGrid = data.frame(degree = 2, scale = 0.1, C = 1.)     
)

yhats = predict(svm.poly, newdata = test)

table(predicted = yhats, actual = test[,1])

get_f1_vals(test[,1],yhats)

stopCluster(cl)
registerDoSEQ()

```

We can see that all the SVMs underperform multiple other models.


## More Feature Selection

```{r}
library(glmnet)
set.seed(1)

x = model.matrix(CrashSeverity ~ ., data = train)[,-1]
y = as.integer(train$CrashSeverity) - 1

cv.lasso = cv.glmnet(x, y, alpha = 1, family = "multinomial")

best.lambda = cv.lasso$lambda.min
lasso.coefs = coef(cv.lasso, s = "lambda.min")

# Obtains the features from lasso that had weights > 0
selected_features <- unique(unlist(lapply(lasso.coefs, function(m) {
  coefs <- as.vector(m)
  names(coefs) <- rownames(m)
  names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"]
})))

# # Fit Lasso with cross-validation
# lasso_cv <- cv.glmnet(x, y, alpha = 1)
# 
# # Coefficients at best lambda
# lasso_coef <- coef(lasso_cv, s = "lambda.min")
# print("Lasso selected features:")
# print(lasso_coef[lasso_coef[,1] != 0, , drop=FALSE])

selected_features = c("CrashSeverity", selected_features)

# Matrix of just the selected features (and target)
data_selected = data_mat[,selected_features, drop = FALSE]

```

## RandomForest

We're using the package ranger as it allows for parallel processing. 


### Random Forest Hyperparameter tuning

```{r p7}
library(randomForest)

split = get_train_test(orig_data,"CrashSeverity", train_pct = .8, data_slice_pct = .025,
                       return_indices = FALSE, seed = 1)
train = split$train
test = split$test

f1s = data.frame(ntrees = numeric(),mtry = numeric(),macro_f1 = numeric(), weighted_f1 = numeric())
trees = seq(50, 500, by = 50)

mtrys = seq(1,13, by = 2)

for (m in mtrys) {
  for (t in trees) {
    print(paste("m:", m,"t:",t))
    
    model.rf = ranger(CrashSeverity ~ ., train, 
                  mtry = m,
                  num.trees = t,
                  num.threads = 16)
    
    yhat = predict(model.rf, test)$predictions

    f1s = rbind(f1s, c(t,m,
                       get_f1_vals(test[,1],yhat)$Macro_F1,
                       get_f1_vals(test[,1],yhat)$Weighted_F1))
  
  }
}


colnames(f1s) = c("ntrees","mtry","macro_f1", "weighted_f1")


```

#### Plotting Hyperparameter tuning

```{r}
par(xpd = TRUE, mar = c(5, 4, 4, 8))  # add extra right margin space
colors = rainbow(length(mtrys))
color_names = c("Red", "Yellow", "Green", "Cyan", "Blue", "Magenta")

f1s = f1s[f1s$mtry != 1,]

plot(NULL,
     xlim = range(f1s$ntrees),
     ylim = range(f1s$macro_f1),
     xlab = "ntrees",
     ylab = "macro f1",
     main = "Macro F1 vs mtrys for different ntree values")

for (i in 1:length(mtrys)) {
  
  subset_data = f1s[f1s$mtry == mtrys[i], ]
  
  lines(subset_data$ntrees, subset_data$macro_f1, col = colors[i], lwd = 2)
}

legend("topright", inset = c(-.3, 0), 
       legend = mtrys,
       col = colors, 
       lwd = 2, 
       title = "MTRY Values", 
       xpd = TRUE, 
       bty = "n")

print("MTRY Values: ")
print(paste(color_names,"=",mtrys))

print(f1s[which.max(f1s$macro_f1),])

plot(NULL,
     xlim = range(f1s$ntrees),
     ylim = range(f1s$weighted_f1),
     xlab = "ntrees",
     ylab = "weighted f1",
     main = "weighted f1 vs mtrys for different ntree values")

for (i in 1:length(mtrys)) {
  
  subset_data = f1s[f1s$mtry == mtrys[i], ]
  
  lines(subset_data$ntrees, subset_data$weighted_f1, col = colors[i], lwd = 2)
}

print("MTRY Values: ")
print(paste(color_names,"=",mtrys))

print(f1s[which.max(f1s$weighted_f1),])


legend("topright", inset = c(-.3, 0), 
       legend = mtrys,
       col = colors, 
       lwd = 2, 
       title = "MTRY Values", 
       xpd = TRUE, 
       bty = "n")



```

### Final RandomForest Model Training

```{r}

split = get_train_test(data,"CrashSeverity", train_pct = .8, data_slice_pct = .1,
                       return_indices = FALSE, seed = 1)
train = split$train
test = split$test


model.rf = ranger(CrashSeverity ~ ., train, 
                  mtry = 5,
                  num.trees = 100,
                  probability = TRUE,
                  num.threads = 16)


yhat_probs = predict(model.rf, test)$predictions

yhats = colnames(yhat_probs)[max.col(yhat_probs, ties.method = "first")]

summary(model.rf)

table(predictions = yhats, actual = test[,1])

get_f1_vals(as.factor(yhats), test[,1])

plot_roc_OvR(yhat_probs, test[,1])


```


## XGBoost

### Creating Label Factors

```{r}

data = orig_data

label_factor <- factor(data[, 1])                 
labels <- as.integer(label_factor) - 1
label_levels <- levels(label_factor)            

```


### Train/test split

```{r}

split = get_train_test(data,"CrashSeverity", train_pct = .5, data_slice_pct = .5, 
                       return_indices = TRUE, seed = 1)

train.inds = split$train
test.inds = split$test

```


### Oversampled Train/test split

```{r}
split = get_train_test(data,"CrashSeverity", train_pct = .99, data_slice_pct = 1,
                       return_indices = TRUE, seed = 1)

test.inds = split$test

split = get_train_test(data[split$train,],"CrashSeverity", train_pct = 1, oversample = TRUE, 
                       return_indices = TRUE, seed = 1)

train.inds = split$train # Oversampled training data


```

### Rest of Train/test splitting

```{r}

train_x = data[train.inds, -1]
train_x = model.matrix(~., train_x)[, -1]

train_y <- labels[train.inds]

# Obtains different split for the test data
split = get_train_test(data[test.inds,],"CrashSeverity", 
                       train_pct = .5,seed = 1, return_indices = TRUE)

valid.inds = split$train

valid_x = data[valid.inds, -1]
valid_x = model.matrix(~., valid_x)[, -1]

valid_y <- labels[valid.inds]

test_x = data[test.inds, -1]
test_x = model.matrix(~., test_x)[, -1]

test_y <- labels[test.inds]

# 4. Create DMatrix
train.xgb <- xgb.DMatrix(data = train_x, label = train_y)

```

### XGBoost Cross Validation

```{r}
params <- list(
  objective = "multi:softprob",
  eval_metric = "mlogloss",
  num_class = length(label_levels),
  eta = .1,
  max_depth = 7,
  min_child_weight = 4,
  gamma = 2,
  nthread = 16
)

cv.model.xgb <- xgb.cv(
  params = params,
  data = train.xgb,
  nfold = 5,
  nrounds = 300,
  early_stopping_rounds = 20,
  verbose = 1,
  prediction = TRUE
)


```

### Classes XGBoost

```{r}
classes.xgb <- xgb.train(
  params = params,
  data = train.xgb,
  nrounds = cv.model.xgb$best_iteration
)

yhats = predict(classes.xgb, test_x)
predicted_labels <- label_levels[yhats + 1]        
actual_labels <- label_levels[test_y + 1]

valid_classes <- intersect(unique(predicted_labels), unique(actual_labels))

# 8. Evaluate
table(predicted = predicted_labels, actual = actual_labels)
get_f1_vals(
  factor(predicted_labels, levels = valid_classes),
  factor(actual_labels, levels = valid_classes)
)
```


### Probabilites XGBoost


```{r}
probs.xgb <- xgb.train(
  params = params,
  data = train.xgb,
  nrounds = cv.model.xgb$best_iteration
)

yhat_probs = predict(probs.xgb, valid_x)
yhat_mat <- matrix(yhat_probs, nrow = length(yhat_probs) / 3, ncol = 3, byrow = TRUE)

yhats <- max.col(yhat_mat) - 1  # -1 if your classes start at 0

yhat_df = as.data.frame(yhat_mat)
colnames(yhat_df) = label_levels

predicted_labels <- label_levels[yhats + 1]        
actual_labels <- label_levels[valid_y + 1]

valid_classes <- intersect(unique(predicted_labels), unique(actual_labels))

# 8. Evaluate
table(predicted = predicted_labels, actual = actual_labels)
get_f1_vals(
  factor(predicted_labels, levels = valid_classes),
  factor(actual_labels, levels = valid_classes)
)

plot_roc_OvR(probs = yhat_df, actual = actual_labels)


```


