---
title: "DPA Model Development"
author: "Jacob Callahan"
date: "2025-04-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyr)
library(caret)
library(nnet)

# Accessing Dataset
csv_data = as.data.frame(read.csv("~/Documents/iit/SP25/DPA/Project/modeling/final_dataset.csv"))

```

## Some data cleaning 

Changes some of the dta

```{r}
unclean_data = csv_data
# Replaces (N/A) with Unable to Determine in Cause1 vector
unclean_data$Cause1 = ifelse(unclean_data$Cause1 == "(N/A)",
                          "Unable to Determine", unclean_data$Cause1)

# Replaces injury in CrashSeverity with injury A,B,C and drops CrashInjurySeverity
# unclean_data$CrashSeverity = ifelse(unclean_data$CrashSeverity == "Injury",
#                             as.character(unclean_data$CrashInjurySeverity), 
#                             as.character(unclean_data$CrashSeverity))

# Drops CrashSeverityCD and CrashInjurySeverity as they have collinearity with CrashSeverity
unclean_data = unclean_data[,-c(1,2)]



# Swapping CrashSeverity to first column
unclean_data = unclean_data[, c("CrashSeverity", "TotalInjured", 
                                setdiff(names(unclean_data), 
                                c("TotalInjured", "CrashSeverity")))]

# This drops TimeOfCrash and DamagedProperty1. These variables need to be fixed to conduct analysis on. 
unclean_data = unclean_data[,!colnames(unclean_data) %in% c("TimeOfCrash","DamagedProperty1","CrashMonth",                            "CityClassCode","CityName","CityClassCode","CrashReportCounty","TotalInjured","TrafficControlDeviceCond")] 

orig_data = unclean_data

```

## Factorizing the categorical variables

```{r}
cols = colnames(orig_data)

num_vars = c("CrashSeverityCd","CityClassCode","TotalInjured","TimeOfCrash","CrashMonth")

for (i in 1:length(orig_data)) {
  if (!colnames(orig_data)[i] %in% num_vars) {
    orig_data[,i] = as.factor(orig_data[,i,drop = TRUE])
  }
}

```

```{r}
data = orig_data[sample.int(nrow(orig_data), size = 1000),]
severity = data$CrashSeverity
```


## Splitting up the data

Here we are splitting the data into training and testing sets. The test set is only 20% of the size of the training data. 

```{r}

n = nrow(data)
train.inds = sample.int(n, size = n*.8)

train = data[train.inds,]
test = data[-train.inds,]

```


## Developing Baseline Model

Source of VGAM package: https://www.geeksforgeeks.org/multinomial-logistic-regression-in-r/#

Our baseline model will be a simple multinomial Logistic Regression. Our response variable will be *CrashSeverity*. Our sole predictor will be *WeatherCond*. The three responses we will be predicting can be Fatal, Injury, or Property Damage.

```{r}


model.log = multinom(CrashSeverity ~ WeatherCond, data = orig_data, family = multinomial)

yhats <- predict(model.log, test, type = "class")
conf_mat = table(predicted = yhats, actual = test$CrashSeverity)
conf_mat

```

Clearly we can see that the baseline model is not very effective at predicting anything but Property Damage. 


## Adding predictors to baseline model

Our initial model was only partially effective as we just had a single predictor. Now we will add all predictors for a multiple multinomial logistic regression.

Using library "nnet" to conduct multinomial logistic regression. 

```{r}
library(nnet)

model.log = multinom(CrashSeverity ~ ., data = train)

yhats = predict(model.log, test, type = 'class')

table(predicted = yhats, actual = test$CrashSeverity)

```


```{r}
cv_model <- function(formula, data, model_func, predict_func, k = 5, ...) {
  set.seed(123)
  folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
  
  all_actuals <- c()
  all_preds <- c()
  
  for (i in 1:k) {
    test_idx <- which(folds == i)
    test_data <- data[test_idx, ]
    train_data <- data[-test_idx, ]
    
    model <- model_func(formula, train_data, ...)
    preds <- predict_func(model, test_data)
    actuals <- model.frame(formula, test_data)[[1]]
    
    all_actuals <- c(all_actuals, as.character(actuals))
    all_preds <- c(all_preds, as.character(preds))
  }

  classes <- sort(unique(c(all_actuals, all_preds)))
  
  actual_counts <- table(factor(all_actuals, levels = classes))
  pred_counts <- table(factor(all_preds, levels = classes))
  
  df <- as.data.frame(t(c(actual_counts, pred_counts)))
  names(df) <- c(paste0("actual_", classes), paste0("pred_", classes))
  
  return(df)
}
```

```{r}
cv_df = cv_model(CrashSeverity ~ ., data, multinom,predict)
cv_df
```

## Basic Plotting

```{r}

# Assume df has 'CrashSeverity' and 'WeatherCond'
df_percent <- csv_data %>%
  group_by(CrashSeverity, DayOfWeek) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(DayOfWeek) %>%
  mutate(percent = 100 * n / sum(n))

df_percent

ggplot(df_percent, aes(x = CrashSeverity, y = percent, fill = DayOfWeek)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Weather Conditions by Crash Severity",
       x = "Crash Severity",
       y = "Percentage") +
  theme_minimal()

df_fatal = df_percent %>% 
  filter(CrashSeverity == "Fatal")

ggplot(df_fatal, aes(x = CrashSeverity, y = percent, fill = DayOfWeek)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Weather Conditions by Crash Severity",
       x = "Crash Severity",
       y = "Percentage") +
  theme_minimal()
```

## RandomForest Analysis

```{r}
library(ranger)
library(MLmetrics)

data = orig_data

train_inds = sample(1:nrow(data), nrow(data) * .8)
  
train = data[train_inds,]
test = data[-train_inds,]
y_test = data[-train_inds,'CrashSeverity']

model.rf = ranger(CrashSeverity ~ ., train, 
                  mtry = 4,
                  num.trees = 500,
                  num.threads = 16)
yhats = predict(model.rf, test, response = 'class')

table(predictions = yhats$predictions, actual = y_test)


f1_per_class <- sapply(levels(y_test), function(class) {
  F1_Score(y_pred = yhats$predictions, y_true = y_test, positive = class)
})

support <- table(test$CrashSeverity)
weighted_f1 <- sum(f1_per_class * support) / sum(support)
weighted_f1
colnames(data)
```
```{r}

```


```{r}
library(ROSE)
balanced_data <- ovun.sample(target ~ ., data = your_data, method = "over", N = 10000)$data
# 
```


