---
title: "DPA Model Development"
author: "Jacob Callahan"
date: "2025-04-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyr)
library(caret)
library(nnet)
source("get_clean_data.R")

orig_data = get_clean_data()
data = orig_data
```

## Model Metric

This function "get_weighted_f1" allows for the weighted f1 based on the test set predictions.

```{r}
get_weighted_f1 = function(y.hats, y.test) {
  f1_per_class <- sapply(levels(y.test), function(class) {
    score <- tryCatch({
      F1_Score(y_pred = y.hats, y_true = y.test, positive = class)
    }, error = function(e) {
      return(0)
    })
    if (is.na(score) || is.nan(score)) score <- 0
    return(score)
  })
  
  support <- table(test$CrashSeverity)
  weighted_f1 <- sum(f1_per_class * support) / sum(support)
  
  if (is.nan(weighted_f1)) {
    print(support)
    print(f1_per_class)
  }
  
  return(weighted_f1)
}
```

# Create one hot encoded matrix for data (Besides target variable)

```{r}
# Creates and saves the labels
label_factor <- factor(orig_data[,1])                 
labels <- as.integer(label_factor) - 1          
label_levels <- levels(label_factor)   

# Turns orig_data into data_mat
# Target variable mapping is 0 - Fatal, 1 - Injury, 2 - Property Damage
data_mat = model.matrix(~., data = orig_data[,-1])
data_mat[,1] = as.integer(orig_data$CrashSeverity) - 1
colnames(data_mat)[1] = "CrashSeverity"

# This is how to turn the target variable back into characters
y = label_levels[data_mat[,1] + 1]
```


## Creating a training and test split

Here we are splitting the data into training and testing sets. The test set is only 20% of the size of the training data. 

```{r}

n = nrow(data)
train.inds = sample.int(n, size = n*.8)

train = data[train.inds,]
test = data[-train.inds,]

```


## Developing Baseline Model

Source of VGAM package: https://www.geeksforgeeks.org/multinomial-logistic-regression-in-r/#

Our baseline model will be a simple multinomial Logistic Regression. Our response variable will be *CrashSeverity*. Our sole predictor will be *WeatherCond*. The three responses we will be predicting can be Fatal, Injury, or Property Damage.

```{r}
model.log = multinom(CrashSeverity ~ WeatherCond, data = orig_data, family = multinomial)

yhats <- predict(model.log, test, type = "class")
conf_mat = table(predicted = yhats, actual = test$CrashSeverity)
conf_mat

```

Clearly we can see that the baseline model is not very effective at predicting anything but Property Damage. 

## Adding predictors to baseline model

Our initial model was only partially effective as we just had a single predictor. Now we will add all predictors for a multiple multinomial logistic regression.

Using library "nnet" to conduct multinomial logistic regression. 

```{r}
library(nnet)

model.log = multinom(CrashSeverity ~ ., data = train)

yhats = predict(model.log, test, type = 'class')

table(predicted = yhats, actual = test$CrashSeverity)

get_weighted_f1(yhats, test[,"CrashSeverity"])
```


# More Feature Selection

```{r}
library(glmnet)
set.seed(1)

sampled_data = orig_data[sample.int(nrow(orig_data), 30000),]

x = model.matrix(CrashSeverity ~ ., data = sampled_data)[,-1]
y = as.integer(sampled_data$CrashSeverity) - 1

cv.lasso = cv.glmnet(x, y, alpha = 1, family = "multinomial")

best.lambda = cv.lasso$lambda.min
lasso.coefs = coef(cv.lasso, s = "lambda.min")

# Obtains the features from lasso that had weights > 0
selected_features <- unique(unlist(lapply(lasso.coefs, function(m) {
  coefs <- as.vector(m)
  names(coefs) <- rownames(m)
  names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"]
})))

selected_features = c("CrashSeverity", selected_features)

# Matrix of just the selected features (and target)
data_selected = data_mat[,selected_features, drop = FALSE]

```

## RandomForest Analysis

We're using the package ranger as it allows for parallel processing. 

```{r}

library(ranger)
library(MLmetrics)

data = orig_data


train = data[train.inds,]
test = data[-train.inds,]
y_test = data[-train.inds,'CrashSeverity']

model.rf = ranger(CrashSeverity ~ ., train, 
                  mtry = 4,
                  num.trees = 500,
                  num.threads = 16)
yhats = predict(model.rf, test, response = 'class')

table(predictions = yhats$predictions, actual = y_test)



get_weighted_f1(yhats$predictions, y_test)


```

## XGBoost

```{r}
library(xgboost)
label_factor <- factor(data[, 1])                 # e.g., "Low", "High", "Medium"
labels <- as.integer(label_factor) - 1            # 0-indexed numeric labels
label_levels <- levels(label_factor)              # store the original meaning


# 2. Extract and encode features
features <- data[, -1]
features_mat <- model.matrix(~., features)[, -1]

# 3. Split into train/test
train_x <- features_mat[train.inds, ]
train_y <- labels[train.inds]

test_x <- features_mat[-train.inds, ]
test_y <- labels[-train.inds]

# 4. Create DMatrix
train.xgb <- xgb.DMatrix(data = train_x, label = train_y)
test.xgb <- xgb.DMatrix(data = test_x)

# 5. Train
params <- list(
  objective = "multi:softmax",
  eval_metric = "mlogloss",
  num_class = length(label_levels),
  nthread = 16
  )

# cv.model.xgb <- xgb.cv(
#   params = params,
#   data = train.xgb,
#   nfold = 5,
#   nrounds = 100,
#   verbose = 1,
#   prediction = TRUE
# )

classes.xgb <- xgb.train(
  params = params,
  data = train.xgb,
  nrounds = 65
)

yhats = predict(classes.xgb, test.xgb)
predicted_labels <- label_levels[yhats + 1]        
actual_labels <- label_levels[test_y + 1]

# 8. Evaluate
table(predicted = predicted_labels, actual = actual_labels)
get_weighted_f1(factor(predicted_labels), factor(actual_labels))




# Matrix of predicted probabilities
yhat_mat <- matrix(predict(final.xgb, test.xgb),ncol = 3, byrow = TRUE)

# Predicted values based on most probable outcome
yhats = max.col(yhat_mat)

# 7. Convert numeric predictions back to original labels
predicted_labels <- label_levels[yhats + 1]        
actual_labels <- label_levels[test_y + 1]

# 8. Evaluate
table(predicted = predicted_labels, actual = actual_labels)
get_weighted_f1(factor(predicted_labels), factor(actual_labels))


```

# Probability analysis to change the outcomes

```{r}
# Predicted Probabilities for fatal when truth != fatal
not_fatal_probs = yhat_mat[test_y != 0,1]

# Predicted Probabilities for fatal when truth == fatal
fatal_probs = yhat_mat[test_y == 0,1]
par(mfrow = c(1,2))
hist(not_fatal_probs, breaks = 1000, 
     xlim = c(0,.1),
     main = "Distribution of Fatal Probablity Predictions when not fatal")
hist(fatal_probs, breaks = 1000, 
     xlim = c(0,.1),
     main = "Distribution of Fatal Probablity Predictions when truly fatal")

threshold = .033
yhats <- apply(yhat_mat, 1, function(row) {
  if (row[1] >= threshold) {
    return(0)
  } else {
    return(which.max(row))
  }
})

sum(yhat_mat[,1] > yhat_mat[,3] & yhat_mat[,1] > yhat_mat[,2]) 

# Quantiles to determine 
quantile(not_fatal_probs, prob = .99)
quantile(fatal_probs, prob = .7)



```

# Binary classification based on fatal

```{r}
library(xgboost)
data = orig_data
data$CrashSeverity = ifelse(data$CrashSeverity == 'Property Damage',"No Injury","Injury")
data$CrashSeverity = as.factor(data$CrashSeverity)

label_factor <- factor(data[,1])                 
labels <- as.integer(label_factor) - 1            
label_levels <- levels(label_factor)              


# 2. Extract and encode features
features <- data[, -1]
features_mat <- model.matrix(~., features)[, -1]


# 3. Split into train/test
train_x <- features_mat[train.inds, ]
train_y <- labels[train.inds]

test_x <- features_mat[-train.inds, ]
test_y <- labels[-train.inds]


# 4. Create DMatrix
train.xgb <- xgb.DMatrix(data = train_x, label = train_y)
test.xgb <- xgb.DMatrix(data = test_x)
dim(test.xgb)
# 5. Train
params <- list(
  booster = "gbtree",
  objective = "binary:logistic",
  eval_metric = "auc",
  nthread = 16,
  scale_pos_weight = sum(train_y == 0) / sum(train_y == 1)
)
# cv.model.xgb <- xgb.cv(
#   params = params,
#   data = train.xgb,
#   nfold = 5,
#   nrounds = 100,
#   verbose = 1,
#   prediction = TRUE
# )

final.xgb <- xgb.train(
  params = params,
  data = train.xgb,
  nrounds = 65
)

# Predicted values based on most probable outcome
yhats_probs = predict(final.xgb, test.xgb)

yhats <- ifelse(yhats_probs > 0.5, 1, 0) 


# 7. Convert numeric predictions back to original labels
predicted_labels <- label_levels[yhats + 1]        
actual_labels <- label_levels[test_y + 1]

# 8. Evaluate
table(predicted = predicted_labels, actual = actual_labels)
get_weighted_f1(factor(predicted_labels), factor(actual_labels))

thresholds <- seq(0.1, 0.9, by = 0.01)
f1_scores <- sapply(thresholds, function(thresh) {
  preds <- ifelse(yhats_probs > thresh, 1, 0)
  F1_Score(y_pred = preds, y_true = test_y, positive = "0")
})
f1_scores

```