---
title: "DPA Model Development"
author: "Jacob Callahan"
date: "2025-04-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(caret)
library(nnet)
library(DBI)
library(RSQLite)
library(dplyr)
library(rsample)
source("utils.R")
library(e1071)
library(xgboost)
library(synthpop)
library(ranger)
library(MLmetrics)
library(doParallel)
```

## Loading Data

```{r}
con <- dbConnect(SQLite(), "crashes.db")

# Get the column names
query <- "
SELECT *
FROM crashes;"
# ORDER BY RANDOM()
# LIMIT 500000;"

orig_data = dbGetQuery(con, query)

# Leaves just CrashSeverity target Column
orig_data = orig_data[,!(colnames(orig_data) %in% c("TotalInjured","CrashInjurySeverity"))]

# Close connection
dbDisconnect(con)

orig_data = orig_data[,!colnames(orig_data) %in% c("HourOfCrash","CrashYr")]
                              
for (col in names(orig_data)) {
  orig_data[[col]] <- factor(orig_data[[col]])
}

data = orig_data

table(data$CrashSeverity)

```

## Developing Baseline Model

Our baseline model will be a simple multinomial Logistic Regression. Our response variable will be *CrashSeverity*. Our sole predictor will be *WeatherCond*. The three responses we will be predicting can be Fatal, Injury, or Property Damage.


### Creating train test split

```{r}

split = get_train_test(data, "CrashSeverity", data_slice_pct = .05, seed = 1)

train = split$train
test = split$test

```

### Baseline Model Training

```{r}
model.log = multinom(CrashSeverity ~ WeatherCond, data = train, family = multinomial)

yhats <- predict(model.log, test, type = "class")
conf_mat = table(predicted = yhats, actual = test$CrashSeverity)
conf_mat

get_f1_vals(test[,1],yhats)

```

Clearly we can see that the baseline model is not very effective at predicting anything but Property Damage. 

## Naive Bayes

```{r}
library(dplyr)
library(e1071)

model.nb = naiveBayes(CrashSeverity ~., data = train, laplace = 0)

yhats = predict(model.nb, test)

table(pred = yhats, actual = test[,1])

get_f1_vals(test[,1],yhats)

```

## Adding predictors to baseline model

Our initial model was only partially effective as we just had a single predictor. Now we will add all predictors for a multiple multinomial logistic regression.

Using library "nnet" to conduct multinomial logistic regression. 

```{r}
library(nnet)

model.log = multinom(CrashSeverity ~ ., data = train)

yhats = predict(model.log, test, type = 'class')

table(predicted = yhats, actual = test$CrashSeverity)

get_f1_vals(test[,1],yhats)

```


## SVM

### Train/Test Split

```{r}

split = get_train_test(orig_data, "CrashSeverity",data_slice_pct = .01)
train = split$train
test = split$test

ctrl <- trainControl(method = "cv", number = 5)  # 5-fold CV

```


### Linear Cross-Validation and Hyperparameter tuning

```{r}
n_cores <- parallel::detectCores()

cl <- makePSOCKcluster(n_cores - 1)  # Leave 1 core free to keep system responsive
registerDoParallel(cl)

grid <- expand.grid(
  cost = c(0.1, 1, 10)        
)

svm.cv.linear <- train(
  CrashSeverity ~ ., 
  data = train,
  trControl = ctrl,
  method = "svmLinear2",
  tuneGrid = grid
)

svm.cv.linear

stopCluster(cl)
registerDoSEQ()

```


Accuracy was used to select the optimal model using the largest value.
The final value used for the model was cost = 0.1.

### Radial Cross-Validation and Hyperparameter tuning

```{r}
n_cores <- parallel::detectCores()

cl <- makePSOCKcluster(n_cores - 1)  # Leave 1 core free to keep system responsive
registerDoParallel(cl)

grid <- expand.grid(
  C = c(0.1, 1, 10),        
  sigma = c(0.0001, 0.001, 0.01)
)

svm.cv.rad <- train(
  CrashSeverity ~ ., 
  data = train,
  trControl = ctrl,
  method = "svmRadial",
  tuneGrid = grid
)

svm.cv.rad

stopCluster(cl)
registerDoSEQ()

```
Accuracy was used to select the optimal model using the largest value.
The final values used for the model were sigma = 0.01 and C = 1

### Radial-Weighted Cross-Validation and Hyperparameter tuning

```{r}
n_cores <- parallel::detectCores()

cl <- makePSOCKcluster(n_cores - 1)  # Leave 1 core free to keep system responsive
registerDoParallel(cl)


grid <- expand.grid(
  C = c(0.1, 1, 10),        
  sigma = c(0.0001, 0.001, 0.01)

)

svm.cv.rad.w <- train(
  CrashSeverity ~ ., 
  data = train,
  trControl = ctrl,
  method = "svmRadialWeights",
  tuneGrid = grid
)

svm.cv.rad.w

stopCluster(cl)
registerDoSEQ()

```


### Polynomial Cross-Validation and Hyperparameter tuning

```{r}

n_cores <- parallel::detectCores()

cl <- makePSOCKcluster(n_cores - 1)  # Leave 1 core free to keep system responsive
registerDoParallel(cl)


grid <- expand.grid(
  degree = c(2, 3, 4),   # degree of the polynomial
  scale = c(0.001, 0.01, 0.1),  # scaling factor
  C = c(0.1, 1, 10)      # regularization parameter
)

svm.cv.poly <- train(
  CrashSeverity ~ ., 
  data = train,
  trControl = ctrl,
  method = "svmPoly",
  tuneGrid = grid
)

svm.cv.poly


stopCluster(cl)
registerDoSEQ()


```

Accuracy was used to select the optimal model using the largest value.
The final values used for the model were degree = 2, scale = 0.1 and C = 1.

### Best Model and prediction

```{r}

yhats = predict(poly.svm.model, newdata = test)

table(predicted = yhats, actual = test[,1])

get_f1_vals(test[,1],yhats)


```






## More Feature Selection

```{r}
library(glmnet)
set.seed(1)

x = model.matrix(CrashSeverity ~ ., data = train)[,-1]
y = as.integer(train$CrashSeverity) - 1

cv.lasso = cv.glmnet(x, y, alpha = 1, family = "multinomial")

best.lambda = cv.lasso$lambda.min
lasso.coefs = coef(cv.lasso, s = "lambda.min")

# Obtains the features from lasso that had weights > 0
selected_features <- unique(unlist(lapply(lasso.coefs, function(m) {
  coefs <- as.vector(m)
  names(coefs) <- rownames(m)
  names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"]
})))

# # Fit Lasso with cross-validation
# lasso_cv <- cv.glmnet(x, y, alpha = 1)
# 
# # Coefficients at best lambda
# lasso_coef <- coef(lasso_cv, s = "lambda.min")
# print("Lasso selected features:")
# print(lasso_coef[lasso_coef[,1] != 0, , drop=FALSE])

selected_features = c("CrashSeverity", selected_features)

# Matrix of just the selected features (and target)
data_selected = data_mat[,selected_features, drop = FALSE]

```

## RandomForest

We're using the package ranger as it allows for parallel processing. 


### Train/Test Split

```{r}
split = get_train_test(data,"CrashSeverity", train_pct = .8, data_slice_pct = .1,
                       return_indices = FALSE, seed = 1)
train = split$train
test = split$test
```


### RandomForest Cross-Validation

```{r}
ctrl <- trainControl(method = "cv", number = 5)

tuneGrid <- expand.grid(
  mtry = c(48, 94), 
  splitrule = c("gini", "extratrees"), 
  min.node.size = c(1, 5, 10)
)

# Correct version with num.threads passed properly:
rf.cv = train(
  CrashSeverity ~ ., 
  data = train, 
  method = "ranger",
  trControl = ctrl,
  tuneGrid = tuneGrid,
  num.threads = 16,         
  importance = "impurity"   
)

print(rf.cv)


```

### Final RandomForest Model Training

```{r}

model.rf = ranger(CrashSeverity ~ ., train, 
                  mtry = 48,
                  num.trees = 250,
                  num.threads = 16)
yhats = predict(model.rf, test, response = 'class')

summary(model.rf)

table(predictions = yhats$predictions, actual = test[,1])

get_f1_vals(yhats$predictions, test[,1])

```


## XGBoost


### Creating Label Factors

```{r}

data = orig_data

label_factor <- factor(data[, 1])                 
labels <- as.integer(label_factor) - 1
label_levels <- levels(label_factor)            

```


### Train/test split with no oversampling

```{r}

split = get_train_test(data,"CrashSeverity", train_pct = .5, data_slice_pct = .025, 
                       return_indices = TRUE, seed = 1)

train.inds = split$train
test.inds = split$test

```


### Train/test split with oversampling

```{r}

split = get_train_test(data,"CrashSeverity", train_pct = .5, oversample = TRUE, 
                       return_indices = TRUE, seed = 1)

train.inds = split$train
test.inds = split$test

```


### Rest of Train/test splitting

```{r}

train_x = data[train.inds, -1]
train_x = model.matrix(~., train_x)[, -1]

train_y <- labels[train.inds]

# Obtains different split for the test data
split = get_train_test(data[test.inds,],"CrashSeverity", 
                       train_pct = .5,seed = 1, return_indices = TRUE)

valid.inds = split$train

valid_x = data[valid.inds, -1]
valid_x = model.matrix(~., valid_x)[, -1]

valid_y <- labels[valid.inds]

test.inds = split$test

test_x = data[test.inds, -1]
test_x = model.matrix(~., test_x)[, -1]

test_y <- labels[test.inds]

# 4. Create DMatrix
train.xgb <- xgb.DMatrix(data = train_x, label = train_y)

```


### Classes XGBoost

```{r}

params <- list(
  objective = "multi:softmax",
  eval_metric = "mlogloss",
  num_class = length(label_levels),
  max_depth = 6,
  min_child_weight = 2,
  gamma = 1,
  nthread = 16
  )

cv.model.xgb <- xgb.cv(
  params = params,
  data = train.xgb,
  nfold = 5,
  nrounds = 70,
  early_stopping_rounds = 20,
  verbose = 1,
  prediction = TRUE
)

classes.xgb <- xgb.train(
  params = params,
  data = train.xgb,
  nrounds = 30,
  cv.model.xgb$best_iteration
)

yhats = predict(classes.xgb, test_x)
predicted_labels <- label_levels[yhats + 1]        
actual_labels <- label_levels[test_y + 1]

valid_classes <- intersect(unique(predicted_labels), unique(actual_labels))

# 8. Evaluate
table(predicted = predicted_labels, actual = actual_labels)
get_f1_vals(
  factor(predicted_labels, levels = valid_classes),
  factor(actual_labels, levels = valid_classes)
)

```


### Probability XGBoost

```{r}
params <- list(
  objective = "multi:softprob",
  eval_metric = "mlogloss",
  num_class = length(label_levels),
  eta = .1,
  max_depth = 7,
  min_child_weight = 4,
  gamma = 2,
  nthread = 16
)

cv.model.xgb <- xgb.cv(
  params = params,
  data = train.xgb,
  nfold = 5,
  nrounds = 300,
  early_stopping_rounds = 20,
  verbose = 1,
  prediction = TRUE
)

probs.xgb <- xgb.train(
  params = params,
  data = train.xgb,
  nrounds = cv.model.xgb$best_iteration
)

yhat_probs = predict(probs.xgb, valid_x)
yhat_mat <- matrix(yhat_probs, nrow = length(yhat_probs) / 3, ncol = 3, byrow = TRUE)

yhats <- max.col(yhat_mat) - 1  # -1 if your classes start at 0

predicted_labels <- label_levels[yhats + 1]        
actual_labels <- label_levels[valid_y + 1]

valid_classes <- intersect(unique(predicted_labels), unique(actual_labels))

# 8. Evaluate
table(predicted = predicted_labels, actual = actual_labels)
get_f1_vals(
  factor(predicted_labels, levels = valid_classes),
  factor(actual_labels, levels = valid_classes)
)


```

